### 2020-07-22 Bayesian Data Analysis/06_Assessing-Model-Assumptions.rmd
bayesian data analysis, ch 6
bayesian data analysis, ch 6

---
title: ''
mainfont: Arial
fontsize: 12pt
documentclass: report
header-includes:
- \PassOptionsToPackage{table}{xcolor}
- \usepackage{caption}
- \usepackage{amssymb}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage[table]{xcolor}
- \usepackage{fancyhdr}
- \usepackage{boldline}
- \usepackage{tipa}
   \definecolor{headergrey}{HTML}{545454}
   \definecolor{msdblue}{HTML}{1C93D1}
   \pagestyle{fancy}
   \setlength\headheight{30pt}
   \rhead{\color{headergrey}\today}
   \fancyhead[L]{\color{headergrey}Moretz, Brandon}
   \fancyhead[C]{\Large\bfseries\color{headergrey}Likelihood}
   \rfoot{\color{headergrey}Chapter 5}
   \lfoot{\color{headergrey}}
   \fancyfoot[C]{\rmfamily\color{headergrey}Bayesian Data Analysis}
geometry: left = 1cm, right = 1cm, top = 2cm, bottom = 3cm
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

@@

library(blmeco, quietly = TRUE, warn.conflicts = FALSE)
library(tidybayes, quietly = TRUE, warn.conflicts = FALSE)
library(arm, quietly = TRUE, warn.conflicts = FALSE)
library(coda, quietly = TRUE, warn.conflicts = FALSE)

@@

# Likelihood

## Theory

Given 3 observations .8, 1.2, 1.1 and a model $y_i \sim Norm(1, .2)$

```{r}
obs <- c(0.8, 1.2, 1.1) # observations

# probability of seeing  these  in n(1, .2)
p1 <- prod(dnorm(obs, mean = 1, sd = 0.2))
p1

# probability of seeing  these  in n(1.2, .4)
p2 <- prod(dnorm(obs, mean = 1.2, sd = 0.4))
p2

p1/p2 # 4x liklier to see under model 1 than model 2

# graphically
xs <- seq(from=0, to=3, by = .01)
ys <- dnorm(xs, mean = 1, sd = .2)

dat <- data.table(x = xs, y = ys)

obs_marks <- data.table(x = obs, y = dnorm(obs, mean = 1, sd = .2))

ggplot(dat) +
   geom_line(aes(x,y), col = "darkgreen", lty=2) +
   geom_hline(yintercept = 0) +
   geom_segment(data = obs_marks, aes(x, y, xend=x, yend=0), col="darkred")

@@

## Quick Look at the Normal Distribution

```{r}
x <- seq(0, 2, length = 100)
dx <- dnorm(x, mean = 1, sd = 0.2)
plot(x, dx, type = "l", xlab = "Weight (kg)", ylab = "Density", lwd = 2, las = 1)

rnorm(5, 1, 0.2) # draws 5 random numbers from Norm(1, 0.2)

pnorm(q = .8, 1, 0.2)
```

$L(\mu, \sigma | y) = \prod^n_{i=1}p(y_i, |\mu, \sigma)$

```{r}
y <- c(0.8, 1.2, 1.1)
lf <- function(mu, sigma) prod(dnorm(y, mu, sigma))

lf(1, .2)
```

## The Maximum Likelihood Method

```{r}
mu <- seq(0.6, 1.4, length=100)
sigma <- seq(0.05, 0.6, length=100)

lik <- matrix(nrow=length(mu), ncol = length(sigma))

for(i in 1:length(mu)) {
   for(j in 1:length(sigma)) {
      lik[i, j] <- lf(mu[i], sigma[j])
   }
}

contour(mu, sigma, lik, nlevels=20, xlab = expression(mu),
        ylab =expression(sigma), las=1, cex.lab = 1.4)

neglf <- function(x) -prod(dnorm(y, x[1], x[2]))
MLest <- optim(c(1, 0.2), neglf)
MLest$par

# optimal values
points(MLest$par[1], MLest$par[2], col="red", pch=4)

@@

Likelihood ratio = p1/p2

## The Log Pointwise Predictive Density

$llpd = \sum^n_{i=1}log\int{p(y_i|\theta)p(\theta|y)d\theta}$

@@

mod <- lm(y~1) # fit model by LS method
nsim <- 2000
bsim <- sim(mod, n.sim=nsim) # simulate from posterior dist. of parameters

pyi <- matrix(nrow=length(y), ncol=nsim)

for(i in 1:nsim) pyi[, i] <- dnorm(y, mean=bsim@coef[i, 1],
                              sd=bsim@sigma[i])

mpyi <- apply(pyi, 1, mean)

sum(log(mpyi))

The log posterior density can be used as a measure of model fit.

>>>>

---
title: ''
mainfont: Arial
fontsize: 12pt
documentclass: report
header-includes:
- \PassOptionsToPackage{table}{xcolor}
- \usepackage{caption}
- \usepackage{amssymb}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage[table]{xcolor}
- \usepackage{fancyhdr}
- \usepackage{boldline}
- \usepackage{tipa}
   \definecolor{headergrey}{HTML}{545454}
   \definecolor{msdblue}{HTML}{1C93D1}
   \pagestyle{fancy}
   \setlength\headheight{30pt}
   \rhead{\color{headergrey}\today}
   \fancyhead[L]{\color{headergrey}Moretz, Brandon}
   \fancyhead[C]{\Large\bfseries\color{headergrey}Assessing Model Assumptions}
   \rfoot{\color{headergrey}Chapter 5}
   \lfoot{\color{headergrey}}
   \fancyfoot[C]{\rmfamily\color{headergrey}Bayesian Data Analysis}
geometry: left = 1cm, right = 1cm, top = 2cm, bottom = 3cm
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

@@

library(blmeco, quietly = TRUE, warn.conflicts = FALSE)
library(tidybayes, quietly = TRUE, warn.conflicts = FALSE)
library(arm, quietly = TRUE, warn.conflicts = FALSE)
library(coda, quietly = TRUE, warn.conflicts = FALSE)
library(nlme, quietly = TRUE, warn.conflicts = FALSE)
library(sp, quietly = TRUE, warn.conflicts = FALSE)
library(gstat, quietly = TRUE, warn.conflicts = FALSE)

@@

# Assessing Model Assumptions

## Model Assumptions

data("periparusater")

dat <- periparusater
mod <- lm(wing ~ sex + age, data = dat)
par(mfrow = c(2, 2))
plot(mod)

@@

```{r}
names(dat)

plot(resid(mod) ~ country, data = dat)
plot(resid(mod) ~ age, data = dat)
plot(resid(mod) ~ sex, data = dat)
plot(resid(mod) ~ weight, data = dat)

@@

```{r}
table(dat$country)

ggplot(dat, aes(y = country)) +
   geom_histogram(aes(fill = ..count..), stat = "count")

## The QQ

```{r}
compareqqnorm(mod)
```

## Temporal Autocorrelation

```{r}
data("parusmajor")

dat <- parusmajor
str(dat)

dat <- dat[order(dat$year, dat$julian), ] # order date acc, time
t.poly.jul <- poly(dat$julian, degree=2)
dat$julian.l <- t.poly.jul[, 1] # orthogonal linear trend
dat$julian.q <- t.poly.jul[, 2] # orthogonal quadratic trend

mod <- lm(count ~ year + julian.l + julian.q, data = dat )

summary(mod)
```

```{r}
par(mfrow = c(1, 2))
acf(resid(mod))
acf(resid(mod), type = "p")
```

### Autocorrelation

```{r}
mod <- gls(count ~ year + julian.l + julian.q, data = dat,
           correlation = corAR1())

summary(mod)
```

### Spatial Autocorrelation

```{r}
data("frogs")

frogs$year.z <- scale(frogs$year)
frogs$elevation.z <- scale(frogs$elevation)
frogs$waterarea[frogs$waterarea == 0 ] <- 0.25
frogs$waterarea.sqrt.l <- log(sqrt(frogs$waterarea))
frogs$waterarea.sqrt.l.z <- scale(frogs$waterarea.sqrt.l)

## Negative Binomial Model

@@

mod <- glm.nb(count2 ~ elevation.z + year.z + fish + vegetation + waterarea.sqrt.l.z + fish:vegetation, data = frogs)

par(mfrow=c(2, 2))
plot(mod)

spdata <- data.frame(resid = resid(mod), x = frogs$x, y = frogs$y)
coordinates(spdata) <- c("x", "y")
bubble(spdata, "resid", col = c("blue", "orange"), main = "Residuals",
       xlab = "X-coordinates", ylab = "Y-coordinates")
```

```{r}
vario.mod <- variogram(resid(mod) ~ 1, spdata)
plot(vario.mod)
```

```{r}
vario.mod.6dir <- variogram(resid(mod) ~ 1, spdata, alpha = seq(0, 150, by = 30))
plot(vario.mod.6dir)
```

## Hetroscedasticity

```{r}
data(ellenberg)
dat <- ellenberg[complete.cases(ellenberg[c("Yi.g", "Water",
                                            "Species")]), ]
mod <- lm(log(Yi.g) ~ Water + Species + Water:Species, dat)
par(mfrow=c(1, 2))
plot(resid(mod) ~ Species, dat)
scatter.smooth(dat$Water, sqrt(abs(resid(mod))), xlab = "Water")
```
